Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-21 papers/contact/paper.tex
rewrite one of the contact function appendices Tested everything on node08.

The procedure for finding the symmetrically averaged correlation
function $g_\sigma^S(\rr)$ is less straight-forward than that of the
previous section, since we must find the average pressure exerted on
the portion of the surface of spheres which is at position $\rr$.  We
achieve this, as in the previous section, by defining a
position-dependent hard-sphere diameter $\sigma(\rr)$, but this time
we define the diameter to be dependent on the position at which the
fundamental measure itself is evaluated:
\begin{align}
  n_3^S(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr)/2)d\rr'
  \label{eq:n3S}
\end{align}
This has the effect of expanding spheres as they touch the point
$\rr$, a concept illustrated in Figure~\ref{fig:contact}.This
enables a similar derivation to that which gave us $g_\sigma^A(\rr)$:
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^S}\\
%  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr)% &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  \sigma(\mathbf{r})}$ in the above formula is computed with the
symmetric formulation of the fundamental measures, as in
Equation~\ref{eq:n3S}.  Details of some aspects of finding this
derivative using FMT are given in Appendix~\ref{appendix:g-S}.  The
equation for $g_\sigma^S$ is computationally simpler than that for
$g_\sigma^A$, as it only involves convolutions of the density itself,
although it does require two new fundamental measures
\begin{align}
  n_{2'}(\rr) &= \int n(\rr') \delta'(\left|\rr - \rr'\right| -
  \sigma/2) d\rr' \\
  \mathbf{n}_{2V'}(\rr) &=
  \int n(\rr') \delta'(\left|\rr - \rr'\right| - \sigma/2) \frac{\rr-\rr'}{|\rr-\rr'|}d\rr'
\end{align}
where $\delta'$ is the derivative of the Dirac delta function.  These
convolutions are easily expressed in reciprocal space, and can be
easily and efficiently computed.

@@

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf
  r''\\
  \frac{\delta n_2^{S}(\rr')}{\delta \sigma(\rr)} &= -\delta(\rr-\rr') \frac 12 \int n(\rr'')
  \delta'(|\rr'-\rr''| - \sigma(\rr')/2) d\rr''
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr') &= \int \frac{n(\rr'')}{2\pi \sigma(\rr')} \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf r''\\
    \frac{\delta n_1^{S}(\rr')}{\delta \sigma(\rr)} &=
    -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi \sigma(\rr)}
    \delta'(|\rr'-\rr''| - \sigma(\rr')/2) d\rr''
    + \frac 12 n_0^s(\rr') \right)
\end{align}
\begin{align}
  n_0^{S}(\rr') &= \int \frac{n(\rr'')}{\pi \sigma(\rr')^2} \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf r''\\
    \frac{\delta n_0^{S}(\rr')}{\delta \sigma(\rr)}
    &= -\delta(\rr-\rr') \left( \frac 12 \int \frac{n(\rr'')}{\pi \sigma(\rr)^2}
    \delta'(|\rr'-\rr''| - \sigma(\rr')/2) d\rr''
    +
    \frac{2}{\sigma} n_0^S(\rr') \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr')/2)
  \frac{(\rr' - \rr'')}{|\rr' - \rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta \sigma(\rr)} &= -\delta(\rr - \rr')
  \frac 12 \int n(\rr'') \delta'(|\rr' - \rr''| - \sigma(\rr')/2)
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr') &= \int \frac{n(\rr'')}{2\pi \sigma(\rr')} \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf r''
   \frac{(\rr' - \rr'')}{|\rr' - \rr''|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta \sigma(\rr)}
 &= -\delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi \sigma(\rr)}
   \delta'(|\rr'-\rr''| - \sigma(\rr')/2) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   + \frac{\mathbf{n}_{V2}^S(\rr')}{2 \pi \sigma(\rr)^2} \right)
\end{align}

The above equations involve just two additional weighted densities
beyond those required for FMT: one scalar density computed with a
derivative of the Dirac delta function, and one analogous vector
density.

>>>>

The procedure for finding the symmetrically averaged correlation
function $g_\sigma^S(\rr)$ is less straight-forward than that of the
previous section, since we must find the average pressure exerted on
the portion of the surface of spheres which is at position $\rr$.  We
achieve this, as in the previous section, by defining a
position-dependent hard-sphere diameter $\sigma(\rr)$, but this time
we define the diameter to be dependent on the position at which the
fundamental measure itself is evaluated:
\begin{align}
  n_3^S(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr)/2)d\rr'
  \label{eq:n3S}
\end{align}
This has the effect of expanding spheres as they touch the point
$\rr$, a concept illustrated in Figure~\ref{fig:contact}.This
enables a similar derivation to that which gave us $g_\sigma^A(\rr)$:
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^S}\\
%  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 2\pi \sigma^2} \frac{\delta
%    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^S(\rr)% &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  \sigma(\mathbf{r})}$ in the above formula is computed with the
symmetric formulation of the fundamental measures, as in
Equation~\ref{eq:n3S}.  Details of some aspects of finding this
derivative using FMT are given in Appendix~\ref{appendix:g-S}.  The
equation for $g_\sigma^S$ is computationally simpler than that for
$g_\sigma^A$, as it only involves convolutions of the density itself,
although it does require two new fundamental measures
\begin{align}
  n_{2'}(\rr) &= \int n(\rr') \delta'(\left|\rr - \rr'\right| -
  \sigma/2) d\rr' \label{n2p} \\
  \mathbf{n}_{2V'}(\rr) &=
  \int n(\rr') \delta'(\left|\rr - \rr'\right| - \sigma/2)
  \frac{\rr-\rr'}{|\rr-\rr'|}d\rr'
  \label{n2vp}
\end{align}
where $\delta'$ is the derivative of the Dirac delta function.  These
convolutions are easily expressed in reciprocal space, and can be
easily and efficiently computed.

@@

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf
  r''\\
  \frac{\delta n_2^{S}(\rr')}{\delta \sigma(\rr)} &= -\delta(\rr-\rr') \frac 12 \int n(\rr'')
  \delta'(|\rr'-\rr''| - \sigma(\rr')/2) d\rr''
  \\
  &= -\delta(\rr-\rr') \frac12 n_{2'}(\rr')
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density, which
we define in Equation~\ref{n2p} as $n_{2'}$.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr') &= \int \frac{n(\rr'')}{2\pi \sigma(\rr')} \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf r''\\
    \frac{\delta n_1^{S}(\rr')}{\delta \sigma(\rr)} &=
    -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi \sigma(\rr)}
    \delta'(|\rr'-\rr''| - \sigma(\rr')/2) d\rr''
    + \frac 12 n_0^s(\rr') \right)
    \\
    &= -\delta(\rr-\rr')\left(
    \frac{n_{2'}(\rr')}{4\pi\sigma} + \frac12n_0(\rr')
    \right)
\end{align}
\begin{align}
  n_0^{S}(\rr') &= \int \frac{n(\rr'')}{\pi \sigma(\rr')^2} \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf r''\\
    \frac{\delta n_0^{S}(\rr')}{\delta \sigma(\rr)}
    &= -\delta(\rr-\rr') \left( \frac 12 \int \frac{n(\rr'')}{\pi \sigma(\rr)^2}
    \delta'(|\rr'-\rr''| - \sigma(\rr')/2) d\rr''
    +
    \frac{2}{\sigma} n_0^S(\rr') \right)
    \\
    &= -\delta(\rr-\rr')\left(
    \frac{n_{2'}(\rr')}{2\pi\sigma^2}
    + \frac{2}{\sigma} n_0(\rr')
    \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr')/2)
  \frac{(\rr' - \rr'')}{|\rr' - \rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta \sigma(\rr)} &= -\delta(\rr - \rr')
  \frac 12 \int n(\rr'') \delta'(|\rr' - \rr''| - \sigma(\rr')/2)
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
  \\
  &= -\delta(\rr-\rr')
   \frac12\mathbf{n}_{V2'}(\rr')
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr') &= \int \frac{n(\rr'')}{2\pi \sigma(\rr')} \delta(|\rr' - \rr''| - \sigma(\rr')/2)d\mathbf r''
   \frac{(\rr' - \rr'')}{|\rr' - \rr''|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta \sigma(\rr)}
 &= -\delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi \sigma(\rr)}
   \delta'(|\rr'-\rr''| - \sigma(\rr')/2) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   + \frac{\mathbf{n}_{V2}^S(\rr')}{2 \pi \sigma(\rr)^2} \right)
   \\
   &= -\delta(\rr-\rr')\left(
   \frac{\mathbf{n}_{V2'}(\rr')}{4\pi \sigma}
   + \frac{\mathbf{n}_{V2}(\rr')}{2 \pi \sigma^2}
   \right)
\end{align}

The above equations involve just two additional weighted densities
beyond those required for FMT: one scalar density $n_{2'}$ computed with a
derivative of the Dirac delta function, and one analogous vector
density $\mathbf{n}_{2V'}$, defined in Equations~\ref{n2p} and \ref{n2vp}.
